---
title: "Posterior Mapping of County Effects with Population Offset"
author: "Jovana Mitic"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
---
We fit a Bayesian negative binomial model with county-specific random effects (theta) and a population offset. Here, we visualize and interpret the posterior means of county effects and their relationship to county population sizes.
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(tigris)
library(rstan)
library(sf)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(patchwork)
library(readr)
library(tidyr)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
sink(tempfile())  
load("data/logbetapop.RData")  # Load fit2 quietly
sink()  

load("data/data_list_august.RData")  # Data list



```

### Mean Posterior Theta Mapping
We extract the posterior means of the random effects, join them with geographic and demographic data, and plot them across South Carolina counties.
### Mean Posterior Theta Mapping

We extract the posterior means of the random effects, join them with geographic and demographic data, and plot them across South Carolina counties.

```{r theta-mapping, fig.width=12, fig.height=6, message=FALSE, warning=FALSE}

# Extract theta samples
# NEW GOOD
theta_samples_with <- as.matrix(fit2, pars = "theta")
mean_theta_with <- colMeans(theta_samples_with)
Y_matrix <- data_list$Y

theta_map_data_with <- data.frame(
  county = rownames(Y_matrix),
  mean_theta = mean_theta_with
) %>%
  mutate(county = trimws(county))

# Load SC counties shapefile
sc_counties_sf <- tigris::counties(state = "SC", cb = TRUE, year = 2020, class = "sf") %>%
  mutate(NAME = trimws(NAME))

# Load and process population data
raw_df <- read_csv("C:\\Users\\jovan\\Box/BoxPHI-PHMR Projects\\Data\\Community Data\\SC County Data\\SC_covid_data_county.csv") %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  group_by(County) %>%
  arrange(County, Date) %>%
  mutate(
    daily_cases = pmax(Cases - lag(Cases, default = first(Cases)), 0),
    daily_deaths = pmax(Deaths - lag(Deaths, default = first(Deaths)), 0)
  ) %>%
  ungroup()

unique_county_population <- raw_df %>%
  group_by(County) %>%
  summarise(population = max(Population, na.rm = TRUE)) %>%
  rename(county = County)

# Merge shapefile and data
sc_counties_merged <- sc_counties_sf %>%
  left_join(unique_county_population, by = c("NAME" = "county")) %>%
  left_join(theta_map_data_with, by = c("NAME" = "county")) %>%
  mutate(mean_theta = ifelse(is.na(mean_theta), 0, mean_theta))

# Save for future use
save(sc_counties_merged, file = 'data/sc_counties_merged.RData')

# Create Population Map
p_population <- ggplot(sc_counties_merged) +
  geom_sf(aes(fill = population), color = "black") +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80") +
  labs(title = "Population by County", fill = "Population") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    legend.position = "right"
  )

# Create Theta Map
p_theta_with <- ggplot(sc_counties_merged) +
  geom_sf(aes(fill = mean_theta), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80", name = "Mean Theta") +
  labs(title = "Mean Posterior Theta (With Population Offset)",
       subtitle = "South Carolina Counties (2020)",
       caption = "Model: logbetapop.RData") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    legend.position = "right"
  )

# Arrange the two maps side by side
gridExtra::grid.arrange(p_population, p_theta_with, ncol = 2)
```

## Posterior Theta Interpretation

The left panel displays the 2020 county-level populations, while the right panel visualizes the posterior mean of county-specific effects (`theta`) from the Bayesian model.

- **Positive theta values** indicate higher-than-expected case counts after adjusting for population.
- **Negative theta values** suggest lower-than-expected outcomes.

Notably:

- Some highly populated counties (e.g., **Richland**, **Charleston**) exhibit positive theta values.
- However, several low-population counties also show substantial deviations, implying that **population size alone does not fully explain spatial variability** in COVID-19 dynamics.

This pattern suggests that **additional unmeasured county-level factors** (such as socioeconomic status, healthcare access, or demographics) may influence the observed outcomes.

---

##Mean Lambda Mapping (Expected Case Counts)

We also visualize the **posterior mean of lambda** (expected number of cases) across months **April–June 2020** to examine the spatial and temporal progression of COVID-19 spread across South Carolina.

```{r lambda-mapping, fig.width=14, fig.height=5, message=FALSE, warning=FALSE}
# Load eta_list if not already loaded
eta_list <- readRDS("data/eta_list.RDS")

# Compute mean and SD for eta
mean_eta <- lapply(eta_list, function(mat) colMeans(mat))
sd_eta <- lapply(eta_list, function(mat) apply(mat, 2, sd))
mean_eta_matrix <- do.call(rbind, mean_eta)
sd_eta_matrix <- do.call(rbind, sd_eta)

# Define dates and months
dates <- as.Date(colnames(data_list$Y))
months_to_plot <- c("2020-04", "2020-05", "2020-06")

# Load SC counties shapefile
sc_counties_sf <- tigris::counties(state = "SC", cb = TRUE, year = 2020, class = "sf") %>%
  mutate(NAME = trimws(NAME))

# Create monthly lambda maps
library(patchwork)
lambda_mean_plots <- list()

for (month in months_to_plot) {
  month_index <- which(format(dates, "%Y-%m") == month)
  
  lambda_map_data <- data.frame(
    county = rownames(data_list$Y),
    mean_lambda = rowMeans(mean_eta_matrix[, month_index, drop = FALSE]),
    std_error = rowMeans(sd_eta_matrix[, month_index, drop = FALSE])
  ) %>%
    mutate(county = trimws(county))
  
  sc_lambda_map <- sc_counties_sf %>%
    left_join(lambda_map_data, by = c("NAME" = "county")) %>%
    mutate(mean_lambda = ifelse(is.na(mean_lambda), 0, mean_lambda))
  
  p <- ggplot(sc_lambda_map) +
    geom_sf(aes(fill = mean_lambda), color = "black", size = 0.2) +
    scale_fill_gradient(low = "blue", high = "red", na.value = "grey50") +
    labs(title = paste("Mean Lambda -", month),
         fill = "Mean Lambda") +
    theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      legend.position = "right"
    )
  
  lambda_mean_plots[[month]] <- p
}

# Combine and display all three months side-by-side
combined_lambda_plot <- wrap_plots(lambda_mean_plots, ncol = 3)
print(combined_lambda_plot)
```


### Discussion: Temporal Dynamics of Expected Cases

The monthly lambda maps reveal the **spatial progression** of COVID-19 activity across South Carolina:

- **April 2020**: Early clusters appeared in **Richland** and **Orangeburg** counties.
- **May 2020**: Emergence of a hotspot in **Greenville** County.
- **June 2020**: Widespread surges observed in **Greenville**, **Horry**, **Charleston**, and **Lexington** counties.

This diffusion pattern suggests **early community spread**, beginning in localized areas and gradually expanding across broader regions of the state.

